{\plant}: A package for modelling forest trait ecology \& evolution

### 2015-05-29 ms.tex
Tidying up many loose ends
closes #19, close #18, closes #24

Check main text for spelling and grammar, further tighten text.
Scientific notation for units in figures.
Running head
Word count
links to other vignettes in demography and physiology documents.
Add details for stem diameter to physiology.

\title{{\plant}: A package for modelling forest trait ecology \& evolution}
\author{Daniel S. Falster\textdagger\textasteriskcentered, Richard G. FitzJohn\textdagger, and Mark Westoby}
\affiliation{
Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia\\
\textdagger These authors contributed equally.\\ 
\textasteriskcentered Email for correspondence: \texttt{daniel.falster@mq.edu.au}\\
A manuscript in consideration as a research paper for
publication in MEE as part of the Special Feature \emph{Demography
  beyond the Population}.}
\date{}

@@

\mstype{Applications Note}
\runninghead{plant}
\keywords{demography, emergent, fitness, growth,
physiology, metapopulation, mortality, reproduction, size-structure,
tradeoff}

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary}\label{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Population dynamics in forests are strongly size-structured:
  larger plants shade smaller plants while also expending
  proportionately more energy on building and maintaining woody stems.
  Although the importance of size-
  structure for demography is widely recognised, many mechanistic models
  either omit it entirely, or include only coarse approximations.
\item
  Here we introduce the {\plant} package, an
  extensible framework for modelling size- and trait-based demography, 
  ecology and evolution in simulated forests.
  At its core, {\plant} is an
  individual-based mechanistic model where plant physiology and demography is mediated by
  traits. Individuals from multiple species can be grown in isolation,
  in patches of competing plants, or in metapopulations under a
  disturbance regime. Dynamics within patches of competing plants are
  resolved using novel extensions of the ``Escalator Boxcar Train''
  technique. Combined effects of trait-, size- and patch-structured
  dynamics are integrated into metapopulation level estimates of
  reproductive fitness and vegetation structure.
\item
  {\plant} is an open source R package and is available at
  \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
  While accessed from R, the core routines in {\plant} are written in C++.
  The package provides for alternative physiologies and for
  hyper-parameterisation to capture trade-offs among parameters. A
  detailed test suite is provided to ensure correct behaviour of the code.
\item
% todo
  We provide worked examples illustrating {\plant}'s use to study the
  influence of functional traits on growth of individual plants, of
  whole patches and on assembly of ecological communities.
\end{enumerate}

@@

Plant growth and demography are fundamentally size- and trait-
structured, influencing dynamics over time-scales ranging from
instantaneous physiological effects to long-term evolutionary outcomes
\citep{Harper-1977, Westoby-2002}. 
%
As an individual increases its leaf area, it increases its potential
to generate photosynthate.
%
In opposition to this, as individuals grow they must use increasing
fractions of their photosynthetic income to building tissues other
than new leaves; to maintaining support tissues \citep{Givnish-1988,
  Enquist-2007} and to reproduction \citep{Thomas-2011}.
%
Consequently rates of growth, mortality and reproduction change with
individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.  The
patterns of changing growth rate vary with plant traits; for example,
leaf construction costs and wood density affect growth rates and rates
of tissue replacement \citep{Wright-2010}, while seed size and height
at maturation determine start and end points of ontogenetic
trajectories \citep{Westoby-2002}.
%
Strong feedbacks emerge between individuals within a forest due to
competition for light and other resources, such that the growth rate
of one individual depends on the size and traits of nearby
individuals.
%
This makes it difficult to predict demographic processes without
modelling both individual growth and the interactions between
individuals.
%
It is therefore difficult to link observable species traits to
phenomena such as self-thinning, successional transitions, trait
evolution and species coexistence.

Although the importance of size-structure for vegetation dynamics and
diversity, has been long been recognised \citep[e.g.,][]{Harper-1977,
  Shugart-1980, Huston-1987}, current research in this area is
dominated by models and theory that either omit size-structure
entirely, or only include coarse approximations. 
%
Theoretical investigations exploring questions about niche-based
species coexistence and ecological drift (neutrality) both primarily
rely on unstructured population models \citep[e.g.][]{MacArthur-1967,
  Tilman-1985, Geritz-1995, Hubbell-2001, Calcagno-2006}.
%
It is common to assume competitive interactions are influenced by
size-related traits such as adult size and seed size, but detailed
size-structure is rarely considered, at least for plant models
\cite[for animal examples, see][]{Deroos-1988, Deroos-1992}.
%
Consequently, it has been difficult to connect and test the mechanisms
described in these simplistic models with types of data collected in
the field, such as physiological leaf traits or demographic data from
long-term monitoring plots \citep{Adler-2013}.
%
Similarly, most large-scale models of vegetation dynamics do not
include size-structured demography \citep{Cramer-2001, Sitch-2003,
  Purves-2008,Dekauwe-2014}.  During the 1980s, size-structured
``gap'' models were widely used
\citep[e.g.][]{Shugart-1980,Huston-1987}. However, the next generation
of models focused more heavily on biogeochemistry at the expense of
demography \citep[e.g.][]{Woodward-1995,Haxeltine-1996,Cramer-2001, 
Sitch-2003}. While the
importance of size-structure is starting again to be recognised
\citep{Moorcroft-2001, Purves-2008, Falster-2011, Dekauwe-2014}, all but a couple
of the most popular,
large-scale vegetation models lack any real demography or
size-structure \citep[e.g.][]{Moorcroft-2001, Smith-2014}.

In this note, we introduce the {\plant} package for R \citep{R-2015};
a mechanistic framework for studying the effects of size structure and
trait variation on the demography of individual plants, of patches of
competing plants and of meta-populations structured by a prevailing
disturbance regime.
% 
We developed {\plant} primarily to investigate if and how species
differing in traits are able to coexist with one another
\citep{Falster-2011,Falster-2015}.  In order to allow interactions to
be driven by species traits and arise through competition for shared
resources, we needed to model a community with a full range of sizes,
as species may not be able to recruit under the canopy that they
generate \citep{Moorcroft-2001,Huston-1987}.

At the same time, we expect the {\plant} model to be useful for
modelling other phenomena. The internal structure of the {\plant}
model is similar in approach to the ``Ecosystem Demography''
\citep{Moorcroft-2001} and \textsc{lpj-guess} \citep{Smith-2014}
models. Like {\plant}, these models include size- structured
competition within a metapopulation of ``patches''. However, whereas
these models were designed to capture large-scale regional
biogeochemistry driven by climatic variables and rely on simulations,
{\plant} was designed so that the rules driving the dynamics of the
system could be manipulated and emergent demographic behaviours
studied deterministically and in detail. In {\plant} it is possible to grow plants in
isolation, in patches of competing plants and in disturbance-driven
metapopulations. 
% TODO: can you change this?  the amount of effort we put in is
% decoupled from how useful other people may find it.
Moreover, we put considerable effort into resolving
competitive hierarchies. 
The package is thus designed to sit in the middle
ground between a large-scale coupled simulation model and simple
abstract theoretical device.

@@

\section{Effects of size, trait and environment on demography of individual plants}Q

The core of {\plant} is a model for an individual plant's
physiological strategy as specified by its traits
(Fig. \ref{fig:schematic}a). This model estimates rate of biomass
production for a plant, given its size and a light environment. Gross
photosynthetic income is estimated from total leaf area and the light
distribution across the plant's canopy. Costs of tissue respiration
and turnover are then subtracted. The remaining biomass is then
allocated between growth and reproduction.  
The key outputs needed by higher levels (see below) is growth rate,
mortality rate and rate of seed production.  However additional
quantities are also computed, such as total assimilation and the
respiration, turnover and allocation rates for different tissues (see
Appendix \ref{sec:plant})

The default physiological model used in {\plant} largely reflects that
presented in \citet{Falster-2011}, but with extensions allowing for
diameter growth also to be estimated. To properly model trait
variation, different parameters (such as rates of leaf turnover,
respiration, photosynthesis per leaf area) in the model are linked via
assumed trade-offs. For example, the trait \textsc{lma} is used to estimate the
rate of leaf turnover, based on observed scaling relationship spanning
across diverse vegetation and plant functional global
\citep{Wright-2004} (see Appendix \ref{sec:FFW16} for details).

The current parametrisation of the physiological model is based on
\citet{Falster-2011}. Parameter values were drawn from available
literature (see Appendix \ref{sec:FFW16}, table 3); however for some
parameters limited or no data were readily available. Providing
updated parameter sets is trivial; the challenge is to infer suitable
values by constraining outputs from {\plant} against diverse data
sources \citep{Lebauer-2012, Keenan-2013}. This is an area in which we
are still actively working and one of the major use cases we envisage
for {\plant}.

While we have implemented one physiological model (\texttt{FFW16}),
the package is designed to allow arbitrary additional physiologies to
be introduced.  That is, rather than varying the parameters of the
equations, the equations themselves can be changed, additional
resources could be competed for and the higher level functionality of
the package can still be used.  As an intermediate level, we allow for
parameters to imply relationships among arbitrary other parameters,
allowing straightforward modelling of trade-offs.  For example,
setting leaf mass per unit area (\textsc{lma}) affects leaf turnover rate,
because it is known that leaves which are cheaper to build have shorter 
lifespan \citep{Wright-2004}.

@@

With the basic physiological model, {\plant} can be used to estimate
essential physiological rates for individual plants
(Fig. \ref{fig:plant}). The function \texttt{grow\_plant\_to\_size}
takes a given strategy and light environment and grows the plant,
producing a trajectory of size over time (Fig. \ref{fig:plant}a). This
is achieved by integrating an Ordinary Differential Equation of
size-dependent growth rate (see Appendices \ref{sec:demography} and
\ref{sec:plant} for more details). As the plant grows, allocation to
different tissue types varies and the composition of the plant
changes to include more stem and less leaf
(Fig. \ref{fig:plant}b). This in turn affects maintenance and turnover
costs and the growth rate of plants (Fig. \ref{fig:plant}c).  This
shift in allocation alone generates a distinctive hump-shaped pattern
of absolute growth with size \citep{King-2011} and equally the widely
regarded decline in relative growth rate with size from birth onwards
\citep{Enquist-2007}.

By varying the light environment and measuring growth rate, the whole
plant light compensation point (\textsc{wplcp}) can be computed (Fig.
\ref{fig:plant}d). The \textsc{wplcp} is the light level at which the plant 
stops growing and is increasingly regarded as the most useful
measure of a strategy's shade tolerance
\citep{Givnish-1988, Baltzer-2007, Lusk-2013}. As expected, \textsc{wplcp}
increases with plant size, due to increased costs of building and
maintaining stem and leaf tissues \citep{Givnish-1988}. Likewise, \textsc{wplcp}
decreases with \textsc{lma}, because high \textsc{lma} species have lower leaf turnover
\citep{Baltzer-2007, Lusk-2013} (Fig.
\ref{fig:plant}d).

@@

Within patches of competing plants, competition for light generates
strong non-linear feedbacks on growth, survival and reproduction. In the
\texttt{FFW16} physiological model, we consider only the effect of
shading on rates of biomass production, although competition for other
resources such as nitrogen or water could also be considered with
suitable extensions.

We assume that patches are vertically but not horizontally
structured. In other words, we account for size differences but not spatial 
layout within the patch. Similar assumptions are found in many models
simulating size-structured dynamics
\citep{Shugart-1980,Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}. In addition, 
we assume that patches are very large -- this means the population dynamics 
within the patch behave deterministically \citep{Deroos-1997} and can therefore 
be approximated via a Partial Differential Equation (\textsc{pde}) (see Appendix
\ref{sec:demography}). The same \textsc{pde} captures average behaviour
across a large number of small patches \citep{Moorcroft-2001}. The model may 
therefore capture average demographic behaviour across a range of patch sizes.

Ideally one would also consider spatial interactions within patches,
however such models are very computationally demanding and also
impossible to solve in a deterministic fashion \citep{Huston-1987, 
Pacala-1996}. Moreover, it remains unclear whether resolving spatial details within a
patch would provide further useful ecological detail, in part because the
process of competitive thinning tends to break down spatial clusters
\citep{Strigul-2008}.

When modelling a patch, we are interested in modelling changes in the size-
density-distribution \(n(x,h,a)\) with patch age \(a\), i.e. the density of
individuals with traits \(x\) and height \(h\) within a patch of age \(a\).
Density is measured as individuals per unit height and per unit ground area.
Modelling the development of  \(n(x,h,a)\) requires that both the initial size
distribution and the inflow of new recruits be specified. We have primarily
been interested in dynamics of a patch recovering from disturbance and have
therefore started with an empty patch and constant flow of seeds from a global
seed rain (Fig. \ref{fig:schematic}b).

@@

Modelling of size-structured dynamics via solving of a deterministic \textsc{pde}
enables multiple complex non-linear demographic phenomena to be
investigated. First, {\plant} can simulate patterns of patch development after
disturbance, including self-thinning behaviours and successional
replacement. Whereas previous approaches for modelling self-thinning
were limited to a single species
\citep[e.g.][]{Barnes-2004, Coomes-2007}, {\plant} accommodates multiple
 species with different traits (Fig. \ref{fig:schematic}b; Fig.
\ref{fig:patch}). Second, {\plant} allows for the effects of competition and succession on
productivity and biomass accumulation within a patch to be investigated
\citep{Falster-2011}. Leaf area cover and rates of biomass production
tend to vary with patch age. Such changes might come about via a
combination of physiological and structural changes
\citep{Binkley-2002, Smith-2001, Coomes-2007, Ogawa-2010}. {\plant} enables
the effects of competitive feedbacks to be incorporated, in addition to
any intrinsic physiological factors. Moreover, {\plant} allows for the
contributions of different species to be mapped (Fig. \ref{fig:patch}c).

@@

In {\plant} we assume an infinite number of patches, all experiencing
the same disturbance regime and linked via seed dispersal (Fig.
\ref{fig:schematic}b). We also assume that the disturbances remove
all established vegetation.  With this approach we can scale up from
simulating a single patch to simulate an entire metapopulation.
The frequency-density
\(p(a)\) of patches aged \(a\) in the landscape is needed. Assuming there
are a large number of patches means the dynamics of \(p\) behave
deterministically and can be approximated via a second \textsc{pde}
(see Appendix \ref{sec:demography} for
details). Scaling from patch to metapopulation is then achieved by
weighting the temporal dynamics of a single patch by \(p(a)\), the
relative abundance of patches aged \(a\) in the metapopulation.

The main numerical challenge in stepping from single patch to
metapopulation is to identify the equilibrium seed rain of the
metapopulation. An ``input'' seed rain is required to simulate 
the demography of the metapopulation, this in turn produces an ``output'' 
number of seeds. The equilibrium seed rain is the vector of seed 
rain where input seed
rain equals output seed rain. Finding this equilibrium requires
finding a root of a multidimensional function, subject to the
constraints that this root is stable (e.g., the trivial equilibrium of
zero seed rain always exists, but is often not stable) (Appendix
\ref{sec:equilibrium}).

A feature of the size-structured metapopulation concept implemented here
is that it balances equilibrium and non-equilibrium approaches to
modelling ecological dynamics \citep{Kohyama-1993, Moorcroft-2001, 
Falster-2011}. An
equilibrium may be approached at the level of metapopulation, meaning
the seed rain and size structure across the entire metapopulation is
approximately stable. Yet the structure of vegetation within individual
patches is constantly in flux. Patches continue to age and their
residents continue to grow until they are disturbed.

@@

The dynamic equilibrium captured by the trait-, size- and patch
structured metapopulation model potentially resolves several
challenges faced by unstructured models. First, by scaling up from
patch to metapopulation, we close the demographic feedback loop to
create a self-regulating system. Unlike some vegetation models, there
is no need to set the density of seeds or seedlings, these arise as
outputs from the model, based on the combined effects of the model's
physiological rules, species traits, competitive interactions and the
prevailing disturbance regime.

Second, the metapopulation concept allows for the effects of disturbance
regimes on vegetation structure and dynamics to be properly incorporated. In
{\plant} characteristics for the entire vegetation (metapopulation) -- such
as total leaf area, biomass and productivity -- are obtained by averaging
patch-level outcomes over the frequency-distribution of patch ages. Because
young patches are more common in the landscape, metapopulation averages are
strongly influenced by the structure of younger patches (e.g. 
Fig. \ref{fig:patch}c).  Similarly, the distributions of  abundance and growth rate
observed in large forest  plots \citep[e.g.][]{Muller-2006, Coomes-2007} arise
as  emergent phenomena. Moreover, the predictions from {\plant} 
are for a frequency distribution of patch states across the landscape 
(Fig. \ref{fig:emergent}, see more detail in Appendix \ref{sec:emergent}). As such,
size-and patch- structured models may be more suitable for linking with 
satellite and forest survey data, than unstructured models representing the 
world as homogeneous green slime \citep{Moorcroft-2001, Purves-2008}.

Finally, with the extension to metapopulation, we are modelling demography across
the entire life cycle and are thus in a position to estimate invasion fitness -- 
the ability of a rare
individual with particular traits to establish in a community of
established residents at their equilibrium densities \citep{Metz-1992}.
Invasion fitness
is ideally calculated as the long-term per capita growth rate of the new
type; however, in structured metapopulations, a more convenient
indicator of per-capita growth rate is given by the logarithm of the
basic reproduction ratio \citep{Gyllenberg-2001, Metz-2001}. The basic
reproduction ratio \(R\) is simply the total seed output of the new
type, averaged across the metapopulation. A new type can invade when
\(R > 1\).

>>>>

\title{{\plant}: A package for modelling forest trait ecology \& evolution}
\author{Daniel S. Falster\textdagger\textasteriskcentered, Richard G. FitzJohn\textdagger, and Mark Westoby}
\affiliation{
Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia\\
\textdagger These authors contributed equally.\\ 
\textasteriskcentered Email for correspondence: \texttt{daniel.falster@mq.edu.au}\\
A manuscript in consideration as a research paper for
publication in MEE as part of the Special Feature \emph{Demography
  beyond the Population}.\\
Word count: ~3900 words}
\date{}

@@

\mstype{Applications Note}
\runninghead{The {\plant} package}
\keywords{demography, emergent, fitness, growth,
physiology, metapopulation, mortality, reproduction, size-structure,
tradeoff}

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary}\label{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Population dynamics in forests are strongly size-structured:
  larger plants shade smaller plants while also expending
  proportionately more energy on building and maintaining woody stems.
  Although the importance of size-
  structure for demography is widely recognised, many mechanistic models
  either omit it entirely, or include only coarse approximations.
\item
  Here we introduce the {\plant} package, an
  extensible framework for modelling size- and trait-based demography, 
  ecology and evolution in simulated forests.
  At its core, {\plant} is an
  individual-based mechanistic model where plant physiology and demography is mediated by
  traits. Individuals from multiple species can be grown in isolation,
  in patches of competing plants, or in metapopulations under a
  disturbance regime. Dynamics within patches of competing plants are
  resolved using novel extensions of the ``Escalator Boxcar Train''
  technique. Combined effects of trait-, size- and patch-structured
  dynamics are integrated into metapopulation level estimates of
  reproductive fitness and vegetation structure.
\item
  {\plant} is an open source R package and is available at
  \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
  While accessed from R, the core routines in {\plant} are written in C++.
  The package provides for alternative physiologies and for
  hyper-parameterisation to capture trade-offs among parameters. A
  detailed test suite is provided to ensure correct behaviour of the code.
\item
  {\plant} provides a transparent platform for investigating how  
  physiological rules and functional trade-offs interact with competition and 
  disturbance regimes to influence vegetation demography, structure and 
  diversity.
\end{enumerate}

@@

Plant growth and demography are fundamentally size- and 
trait-structured, influencing dynamics over time-scales ranging from
instantaneous physiological effects to long-term evolutionary outcomes
\citep{Harper-1977, Westoby-2002}. 
%
As an individual increases its leaf area, it increases its potential
to generate photosynthate.
%
In opposition to this, as individuals grow they must use increasing
fractions of their photosynthetic income to building tissues other
than new leaves; to maintaining support tissues \citep{Givnish-1988,
  Enquist-2007} and to reproduction \citep{Thomas-2011}.
%
Consequently rates of growth, mortality and reproduction change with
individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.  
Ontogenetic 
patterns of growth also vary with traits; for example,
leaf and wood construction costs influence growth rates 
\citep{Wright-2010}, while seed size and height
at maturation determine the start and end points of ontogenetic
trajectories \citep{Westoby-2002}.
%
Strong feedbacks emerge between individuals within a forest via
competition for light and other resources, such that the growth rate
of one individual depends on the size and traits of nearby
individuals \cite{Pacala-1996}.
%
Such feedbacks make it difficult to link observable species traits to
phenomena such as self-thinning, successional transitions, trait
evolution and species coexistence, without
modelling both individual growth and the competitive interactions between
individuals.

Although the importance of size-structure for vegetation dynamics and
diversity, has been long been recognised \citep[e.g.,][]{Harper-1977,
  Shugart-1980, Huston-1987}, current research in this area is
dominated by models and theory that either omit size-structure
entirely, or only include coarse approximations. 
%
Theoretical investigations exploring questions about niche-based
species coexistence and ecological drift (neutrality) both rely
primarily on unstructured population models \citep[e.g.][]{MacArthur-1967,
  Tilman-1985, Geritz-1995, Hubbell-2001, Calcagno-2006}.
%
It is common to assume competitive outcomes are influenced by
size-related traits such as adult size and seed size, but detailed
size-structure is rarely considered, at least in plant models
\cite[for animal examples, see][]{Deroos-1988, Deroos-1992}.
%
Similarly, most large-scale models of vegetation dynamics do not
include size-structured demography \citep{Cramer-2001, Sitch-2003,
  Purves-2008,Dekauwe-2014}.  During the 1980s, size-structured
``gap'' models were widely used
\citep[e.g.][]{Shugart-1980,Huston-1987}. However, the next generation
of models focused more heavily on biogeochemistry at the expense of
demography \citep[e.g.][]{Woodward-1995,Haxeltine-1996,Cramer-2001, 
Sitch-2003}. While the
importance of size-structure is starting again to be recognised
\citep{Moorcroft-2001, Purves-2008, Falster-2011, Dekauwe-2014}, all but a couple
of the most popular,
large-scale vegetation models lack any real demography or
size-structure \citep[see][]{Moorcroft-2001, Smith-2014}.

In this note, we introduce the {\plant} package for R \citep{R-2015};
a mechanistic framework for studying the effects of size structure and
trait variation on the demography of individual plants, of patches of
competing plants and of meta-populations structured by a prevailing
disturbance regime.
% 
We developed {\plant} primarily to further investigate if and how species
differing in traits are able to coexist with one another
\citep[following][]{Falster-2011,Falster-2015}.  Size-structured models
provide unique opportunities for coexistence via differentiation in
successional strategy \citep{Moorcroft-2001,Huston-1987}.

At the same time, we expect the {\plant} model to be useful for
modelling other demographic phenomena. The internal structure of the {\plant}
model is similar in approach to the ``Ecosystem Demography''
\citep{Moorcroft-2001} and \textsc{lpj-guess} \citep{Smith-2014}
models. Like {\plant}, these models include size-structured
competition within a metapopulation of ``patches''. However, whereas
these models rely on simulations and were designed to capture large-scale regional
biogeochemistry driven by climatic variables,
{\plant} was designed so that the rules driving the dynamics of the
system could be manipulated and emergent demographic behaviours
studied deterministically and in detail. In {\plant} it is possible to grow plants in
isolation, in patches of competing plants and in disturbance-driven
metapopulations. The package is thus designed to bridge the gap between 
a large-scale coupled simulation model and simple
abstract theoretical device.

@@

\section{Effects of size, trait and environment on demography of individual plants}

The core of {\plant} is a model for an individual plant's
physiological strategy as specified by its traits
(Fig. \ref{fig:schematic}a). This model estimates rate of biomass
production for a plant, given its size and a light environment. Gross
photosynthetic income is calculated from total leaf area and the light
distribution across the plant's canopy. Costs of tissue respiration
and turnover are then subtracted. The remaining biomass is then
allocated between growth and reproduction.  
The key outputs needed by higher levels (see below) are height growth rate,
mortality rate and rate of seed production.  However additional
quantities are also computed, such as total assimilation and the
respiration, turnover and allocation rates for different tissues (see
Appendix \ref{sec:plant})

The default physiological model used in {\plant} largely reflects that
presented in \citet{Falster-2011,Falster-2015}, but with extensions allowing for
diameter growth also to be estimated. To properly model trait
variation, different parameters (such as rates of leaf turnover,
respiration, photosynthesis per leaf area) in the model are linked via
assumed trade-offs.

The current parametrisation of the physiological model is based on
\citet{Falster-2011}. Parameter values were drawn from available
literature (see Table 3 in Appendix \ref{sec:FFW16}); however for some
parameters limited or no data were readily available. Providing
updated parameter sets is trivial (see  Appendix \ref{sec:parameters}); 
the challenge is to infer suitable
values by constraining outputs from {\plant} against diverse data
sources \citep{Lebauer-2012, Keenan-2013}. This is an area in which we
are still actively working and one of the major use cases we envisage
for {\plant}.

While we have implemented one physiological model (\texttt{FFW16}),
the package is designed to allow arbitrary additional physiologies to
be introduced.  That is, rather than varying the parameters of the
equations, the equations themselves can be changed and the higher level 
functionality of the package can still be used.  As an intermediate level, 
we allow for changes in key parameters to trigger changes 
in other parameters (i.e. hyper-parametrisation),
allowing straightforward modelling of trade-offs.  
For example, the trait \textsc{lma} is used to estimate the
rate of leaf turnover, based on an observed scaling relationship spanning
across diverse vegetation and plant functional types
\citep{Wright-2004} (see Appendix \ref{sec:FFW16} for details).

@@

With the basic physiological model, {\plant} can be used to estimate
essential physiological rates for individual plants
(Fig. \ref{fig:plant}). The function \texttt{grow\_plant\_to\_size}
takes a given strategy and light environment and grows the plant,
producing a trajectory of size over time (Fig. \ref{fig:plant}a). This
is achieved by integrating an ordinary differential equation of
size-dependent growth rate (see Appendices \ref{sec:demography} and
\ref{sec:plant} for more details). As the plant grows, allocation to
different tissue types varies and the composition of the plant
changes to include more stem and less leaf
(Fig. \ref{fig:plant}b). This in turn affects maintenance and turnover
costs and the growth rate of plants (Fig. \ref{fig:plant}c).  This
shift in allocation alone generates a distinctive hump-shaped pattern
of absolute growth with size \citep{King-2011} and equally the widely
regarded decline in relative growth rate with size from birth onwards
\citep{Enquist-2007}.

By varying the light environment and measuring growth rate, the whole
plant light compensation point (\textsc{wplcp}) can be computed (Fig.
\ref{fig:plant}d). The \textsc{wplcp} is the light level at which the plant 
stops growing and is increasingly regarded as the most useful
measure of a species's shade tolerance
\citep{Givnish-1988, Baltzer-2007, Lusk-2013}. As expected, \textsc{wplcp}
increases with plant size, due to increased costs of building and
maintaining stem and leaf tissues \citep{Givnish-1988}. Likewise, \textsc{wplcp}
decreases with \textsc{lma}, because high \textsc{lma} species have lower leaf turnover
\citep{Baltzer-2007, Lusk-2013} (Fig.
\ref{fig:plant}d).

@@

Within patches of competing plants, competition for light generates
strong non-linear feedbacks on growth, survival and reproduction. In the
\texttt{FFW16} physiological model, we consider only the effect of
shading on rates of biomass production, although competition for other
resources such as nitrogen or water could also be considered by extending the 
physiological model.

We assume that patches are vertically but not horizontally
structured. In other words, we account for size differences but not spatial 
layout within the patch. Similar assumptions are found in many models
simulating size-structured dynamics
\citep{Shugart-1980,Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}. In addition, 
we assume that patches are very large -- this means the population dynamics 
within the patch behave deterministically \citep{Deroos-1997} and can therefore 
be approximated via a Partial Differential Equation (\textsc{pde}) (see Appendix
\ref{sec:demography}). The same \textsc{pde} captures average behaviour
across a large number of small patches \citep{Moorcroft-2001}. {\plant} may 
therefore capture average demographic behaviour across a range of patch sizes.

Ideally one would also consider spatial interactions within patches,
however such models are very computationally demanding and also
impossible to solve in a deterministic fashion \citep{Shugart-1980, 
Pacala-1996}. Moreover, it remains unclear whether resolving spatial details within a
patch would provide further useful ecological detail, in part because the
process of competitive thinning tends to break down spatial clusters
\citep{Strigul-2008}.

When modelling a patch, the focus is on changes to the size-
density-distribution \(n(x,h,a)\) with patch age \(a\), i.e. the density of
individuals with traits \(x\) and height \(h\) within a patch of age \(a\).
Density is measured as individuals per unit height and per unit ground area.
Modelling the development of  \(n(x,h,a)\) requires that both the initial size
distribution and the inflow of new recruits be specified. We have primarily
been interested in dynamics of a patch recovering from disturbance and have
therefore started with an empty patch and constant flow of seeds from a global
seed rain (Fig. \ref{fig:schematic}b).

@@

Modelling of size-structured dynamics via solving of a deterministic \textsc{pde}
enables multiple complex non-linear demographic phenomena to be
investigated. First, {\plant} can simulate patterns of patch development after
disturbance, including self-thinning behaviours and successional
replacement. Whereas previous approaches for modelling self-thinning
were limited to a single species
\citep[e.g.][]{Barnes-2004, Coomes-2007}, {\plant} accommodates multiple
 species with different traits (Fig. \ref{fig:schematic}b; Fig.
\ref{fig:patch}). Second, {\plant} allows for the effects of competition and succession on
productivity and biomass accumulation within a patch to be investigated
\citep{Falster-2011}. Leaf area cover and rates of biomass production
tend to vary with patch age 
\citep{Binkley-2002, Smith-2001, Coomes-2007, Ogawa-2010}. {\plant} enables
the effects of competitive feedbacks to be incorporated, in addition to
any intrinsic physiological factors, yielding emergent non-linear outcomes
 (Fig. \ref{fig:patch}c).

@@

In {\plant} we assume an infinite number of patches, all experiencing
the same disturbance regime and linked via seed dispersal (Fig.
\ref{fig:schematic}b). We also assume that the disturbances remove
all established vegetation.  With this approach we can scale up from
simulating a single patch to simulate an entire metapopulation.
The frequency-density
\(p(a)\) of patches aged \(a\) in the landscape is needed. Assuming there
are a large number of patches means the dynamics of \(p\) behave
deterministically and can be approximated via a second \textsc{pde}
(see Appendix \ref{sec:demography} for
details). Scaling from patch to metapopulation is then achieved by
weighting the temporal dynamics of a single patch by \(p(a)\), the
relative abundance of patches aged \(a\) in the metapopulation 
 (Fig. \ref{fig:patch}c).

The main numerical challenge in stepping from single patch to
metapopulation is to identify the equilibrium seed rain of the
metapopulation. An ``input'' seed rain is required to simulate 
the demography of the metapopulation, this in turn produces an ``output'' 
number of seeds. The equilibrium occurs when a given
input seed
rain produces an equivalent output seed rain. Finding this equilibrium requires
finding the root of a multi-dimensional function, subject to the
constraints that this root is stable (e.g., the trivial equilibrium of
zero seed rain always exists, but is often not stable) (Appendix
\ref{sec:equilibrium}).

An attractive  feature of the metapopulation concept implemented here
is that integrates equilibrium and non-equilibrium approaches to
modelling ecological dynamics \citep{Kohyama-1993, Moorcroft-2001, 
Falster-2011}. An
equilibrium may be approached at the level of metapopulation, meaning
the seed rain and size structure across the entire metapopulation is
approximately stable. Yet the structure of vegetation within individual
patches is constantly in flux. Patches continue to age and their
residents continue to grow until they are disturbed.

@@

The dynamic equilibrium captured by the trait-, size- and 
patch-structured metapopulation model potentially resolves several
challenges faced by unstructured models. First, by scaling up from
patch to metapopulation, we close the demographic feedback loop to
create a self-regulating system. Unlike some vegetation models, there
is no need to set the density of seeds or seedlings, these arise as
outputs from the model, based on the combined effects of the model's
physiological rules, species traits, competitive interactions and the
prevailing disturbance regime.

Second, the metapopulation concept allows for the effects of disturbance
regimes on vegetation structure and dynamics to be properly incorporated. In
{\plant} characteristics for the entire vegetation (metapopulation) -- such
as total leaf area, biomass and productivity -- are obtained by averaging
patch-level outcomes over the frequency-distribution of patch ages. Because
young patches are more common in the landscape, metapopulation averages are
strongly influenced by the structure of younger patches (e.g. 
Fig. \ref{fig:patch}c).  Similarly, the distributions of  abundance and growth rate
observed in large forest  plots \citep[e.g.][]{Muller-2006, Coomes-2007} arise
as  emergent phenomena. Moreover, the predictions from {\plant} 
are for a frequency distribution of patch states across the landscape 
(Fig. \ref{fig:emergent}, see more detail in Appendix \ref{sec:emergent}). As such,
size- and patch-structured models provide unique opportunities for linking with 
satellite and forest survey data \citep{Moorcroft-2001, Purves-2008}.

Finally, with the extension to metapopulation, we are modelling demography across
the entire life cycle and are thus in a position to estimate invasion fitness -- 
the ability of a rare
individual with particular traits to establish in a community of
established residents at their equilibrium densities \citep{Geritz-1998,Dieckmann-2007}.
While ideally calculated as the long-term per capita growth rate, 
a more convenient
indicator of invasion fitness in structured metapopulations is given by the logarithm of the
basic reproduction ratio, \(R\) \citep{Gyllenberg-2001, Metz-2001}. 
\(R\) is simply the total seed output of the new
type, averaged across the metapopulation. So a new type can invade when
 \(\log R > 0\).

With fitness defined, it becomes possible to model trait-based adaptive dynamics
and community assembly \citep{Falster-2015}. This is best
thought of as occurring on a fitness landscape, a surface with fitness
plotted against one or more species traits (Fig. \ref{fig:fitness}). The
reproductive success of individuals possessing a particular trait value
depends on both the traits and densities of other species in the current
resident community. Fitness landscapes will be positive in regions that allow invasion,
and zero for species that are at equilibrium. The slope of the fitness
landscape at a resident indicates the direction of selection; a
negative slope (dashed line in Fig. \ref{fig:fitness}a) indicates
selection for smaller trait values (see Appendix \ref{sec:fitness} for
details). Fitness landscapes are constantly reshaped as species are
introduced and evolve, reflecting the non-linear demographic feedbacks
that occur under frequency-dependent evolution. Stale coexistence is possible when 
species with different traits can mutually invade \citep{Geritz-1998}
Following the fitness gradient,
can lead to an evolutionary branching point
\citep{Geritz-1998,Dieckmann-2007}; at this point there is no
directional selection (solid line in Fig. \ref{fig:fitness}a) but
invasion is possible both above and below it in trait space. It is
often possible to find a configuration where fitness is negative
everywhere (solid line in Fig. \ref{fig:fitness}b) except for the zero
points where species are present. This community is immune to invasion
from any other strategies and represents a prediction for the 
trait mixture that might might be favoured by natural selection 
\citep{Geritz-1998,Falster-2015}.
